A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-10 tex/sec5-synthesis.tex
[ tex ] Fix the claimproof environment
[ tex ] Revise S5 further

Now let us define mode-correctness rigorously.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For an argument list $\biargvec$,\LT{Eww...} define the set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
    \label{eq:synvar2}\synvar(\biargvec, \chkbiarg[n]) & = \phantom{\fv(A_{n}) \cup {}} \synvar(\biargvec) \\
    \label{eq:synvar3}\synvar(\biargvec, \synbiarg[n]) & = \fv(A_{n}) \cup           \synvar(\biargvec).
  \end{align}
\end{definition}
This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.

@@

Now, we set out to show the uniqueness of the synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction.
However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar\left(\biargs\right)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for~\eqref{eq:synvar2} by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left(\biargs\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
      \end{enumerate}
  \end{itemize}
\end{proof}

@@

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set\/ $\Xi$ of variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{definition}
  By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and\/ $\sigma(x) = \rho(x)$ for every\/ $x$ in the domain of $\rho$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{definition}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or 
    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for the decidability proof.

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed above.
  (For brevity we omit the subscript $(\Sigma, \Omega)$.)
  \begin{itemize}
    \item \SynRule{Var}: As $\erase{\Gamma} |- \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.

@@

    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |- t^{\syn}$ by inversion.
      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{itemize}
        \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
          \begin{itemize}
            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
          \end{itemize}
      \end{itemize}
    \item \Rule{Op}:
      Consider the mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$.
      First we claim:
      \begin{claim}\label{lem:args-induction}
        For an argument list $\as = \biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
        \begin{enumerate}
          \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
            \begin{equation} \label{eq:claim}
              \text{$\ext{\rho}$ is defined on $\synvar(\as)$ and\/} \quad
              \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
              \quad\text{for $i = 1, \ldots, n$};
            \end{equation}
          \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
        \end{enumerate}
      \end{claim}
      Then, we proceed with a case analysis on $\dir{d}$:
      \begin{itemize}
        \item $\dir{d}$ is $\syn$: We apply our claim with the partial substitution $\rho_0$ defined nowhere.
          \begin{enumerate}
            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but $\Gamma |- \tmOpts \syn A$ for some $A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that
              \[
                \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}
              \]
              for every $i$.
              Obviously, $\rho \geq \rho_0$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.

@@

                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined and by the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
                    \begin{enumerate}
                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |/- \isTerm{t_{m+1}} \syn A$ for any $A$ but there is $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m+1$, then $\gamma \geq \ext{\rho}$.
                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify} we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
                        \begin{itemize}
                          \item Suppose no $\sigma \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ such that \eqref{eq:claim} holds for $1 \leq i \leq m + 1$. 
                            Then, $\gamma \geq \ext{\rho}$ by the minimality of $\ext{\rho}$ and thus
                            $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
                          \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
                        \end{itemize}
                    \end{enumerate}
                \end{itemize}
            \end{enumerate}
        \end{enumerate}
        \hfill{\footnotesize$\blacksquare$}
      \end{claimproof}
      We now have completed the decidability proof.
  \end{itemize}
\end{proof}

The formal counterpart of the above proof in \Agda essentially functions as two programs for type checking and synthesis.
These programs either compute the typing derivation or provide a proof of contradiction.
Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
Alternatively, we could consider generalising typing derivations like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
This could accommodate ill-typed cases and improve contradiction proofs to deliver more informative error messages.
This would assist users in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.

>>>>

Now let us define mode-correctness rigorously.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For an argument list $\biargs$,\LT{Eww...} define its set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align}
\end{definition}
This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.

@@

Now, we set out to show the uniqueness of the synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction.
However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for~\eqref{eq:synvar2} by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
      \end{enumerate}
  \end{itemize}
\end{proof}

@@

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set\/ $\Xi$ of variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{definition}
By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{definition}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or 
    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for the decidability proof.

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed above.
  For brevity we omit the subscript $(\Sigma, \Omega)$.
  \begin{itemize}
    \item \SynRule{Var}: As $\erase{\Gamma} |- \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.

@@

    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |- t^{\syn}$ by inversion.
      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{itemize}
        \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
          \begin{itemize}
            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
          \end{itemize}
      \end{itemize}
    \item \Rule{Op}:
      For a mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$, we first claim:
      \begin{claim}\label{lem:args-induction}
        For an argument list $\biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
        \begin{enumerate}
          \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
            \begin{equation} \label{eq:claim}
              \dom(\ext{\rho}) \supseteq \synvar(\biargs)
              \quad\text{and}\quad
%              \text{the domain of $\ext{\rho}$ contains $\synvar(\biargs)$ and\/} \quad
              \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
              \quad\text{for $i = 1, \ldots, n$};
            \end{equation}
          \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
        \end{enumerate}
      \end{claim}

      Then, we proceed with a case analysis on $\dir{d}$ in the mode derivation:
      \begin{itemize}
        \item $\dir{d}$ is $\syn$: We apply our claim with the partial substitution $\rho_0$ defined nowhere.
          \begin{enumerate}
            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but $\Gamma |- \tmOpts \syn A$ for some $A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that
              \[
                \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}
              \]
              for every $i$.
              Obviously, $\rho \geq \rho_0$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.

@@

                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined and by the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
                    \begin{enumerate}
                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |/- \isTerm{t_{m+1}} \syn A$ for any $A$ but there is $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m+1$, then $\gamma \geq \ext{\rho}$.
                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify} we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
                        \begin{itemize}
                          \item Suppose no $\sigma \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ such that \eqref{eq:claim} holds for $1 \leq i \leq m + 1$. 
                            Then, $\gamma \geq \ext{\rho}$ by the minimality of $\ext{\rho}$ and thus
                            $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
                          \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
                        \end{itemize}
                    \end{enumerate}
                \end{itemize}
            \end{enumerate}
        \end{enumerate}
        Therefore, we have proved our claim for any argument list.
      \end{claimproof}
  \end{itemize}
  We now have completed the decidability proof.
\end{proof}

The formal counterpart of the above proof in \Agda functions as two top-level programs for type checking and synthesis.
These programs either compute the typing derivation or provide a proof of contradiction.
Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
Alternatively, we could consider generalising typing derivations like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
This could accommodate ill-typed cases and improve contradiction proofs to deliver more informative error messages.
This would assist users in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.
